Making multiple-locus association mapping on a genome-wide scale routine

### 2017-11-02 AM+draft.tex
02Nov17 end of day updates - methods

Over the past decade,  Genome wide association studies have changed considerably in both their analysis and design. Early gwas followed a case-control design. Association mapping methods were no more complicated than contingency table tests or regression. These designs though had a tendency to yield spurious findings if there was unrecognised population stratification. This prompted a shift towards family-based designs and score tests, such as the tdt test and its variants (refs). Today, instead of by design, it is through statistical modelling that we account for the effects of population stratification. This has meant that data can be collected from general populations, even if these populations are highly structured. Analysis via sophisticated association mapping methods, such as linear mixed model based approaches,  is now almost routine.

@@

\textbf{glmnet:}  Glmnet is a general purpose R package for regression analysis  \cite{Friedman2010glmnet}. 
The model used for analysis is 
$y = X \beta + \sum_{l=1}^L m_l \theta_l + e$ where $\theta_l; l=(1,2, \ldots, L)$ is the coefficient for the $l$th snp. 
All snp effects are modelled, simultaneously. 
Glmnet implements elastic net for the estimation of the parameters. Elastic net is a regularisation method where the 
penalty function is a mixture of 
$\ell_1$-norm (lasso) and $\ell_2$-norm (ridge regression) penalties  \cite{zou2005regularization}. 
This allows elastic net to avoid the limitations of lasso 
while still being able to perform shrinkage and variable selection, simultaneously \cite{zou2005regularization}. 
Glmnet was not specifically built for association mapping. 
However, it is included in our list of comparison applications due to its solid performance for genomic prediction 
\cite{heslot2012genomic,ogutu2012genomic,}.
The glmnet package 
implements a modification of the LARS algorithm \cite{efron2004least} for finding the optimal regularisation (solution) path. 
From exploratory analyses of the simulated data, we found this algorithm yielded a high proportion of false positives. 
We instead fixed the shrinkage parameter to a suitable value 
and employed stability selection to estimate the significance of the snp coefficients, empirically (see below).

@@

  r2VIM

Single locus methods

>>>>

Over the past decade,  Genome wide association studies have changed considerably in both their analysis and design. Early gwas followed a case-control design. Association mapping methods were no more complicated than contingency table tests or simple 
linear regression. These designs though had a tendency to yield spurious findings if there was unrecognised population stratification. This prompted a shift towards family-based designs and score tests, such as the tdt test and its variants (refs). Today, instead of by design, it is through statistical modelling that we account for the effects of population stratification. This has meant that data can be collected from general populations, even if these populations are highly structured. Analysis via sophisticated association mapping methods, such as linear mixed model based approaches,  is now almost routine.

@@

\textbf{glmnet:}  Glmnet is a general purpose R package for regression analysis  \cite{Friedman2010glmnet}. 
The model used for analysis is 
$y = X \beta + \sum_{l=1}^L m_l \theta_l + e$ where $\theta_l; l=(1,2, \ldots, L)$ is the coefficient for the $l$th snp. 
All snp effects are modelled, simultaneously. 
Glmnet implements elastic net for the estimation of the parameters. Elastic net is a regularisation method where the 
penalty function is a mixture of 
$\ell_1$-norm (lasso) and $\ell_2$-norm (ridge regression) penalties  \cite{zou2005regularization}. 
This allows elastic net to avoid the limitations of lasso 
while still being able to perform shrinkage and variable selection, simultaneously \cite{zou2005regularization}. 
Glmnet was not specifically built for association mapping. 
However, it is included in our list of comparison applications due to its solid performance for genomic prediction 
\cite{heslot2012genomic,ogutu2012genomic}.
The glmnet package 
implements a modification of the LARS algorithm \cite{efron2004least} for finding the optimal regularisation (solution) path. 
From exploratory analyses of the simulated data, we found this algorithm yielded a high proportion of false positives. 
We instead fixed the shrinkage parameter to a suitable value 
and employed stability selection to estimate the significance of the snp coefficients, empirically (see below).

@@

\textbf{r2VIM:} r2VIM is an R package, specifically designed for genome-wide association mapping \cite{szymczak2016r2vim}.
It implements a modified form of random forests. When random forests are applied to association study data in a standard way, 
the result is a ranking of the snp according to their importance values. The difficulty is in knowing how many of the top ranked 
snp are associated, significantly, with the trait. In r2VIM, multiple separate random forest analyses of the data are performed.

In r2VIM, a strategy is implemented for identifying the snp in strongest association with the trait. 
Here, multiple random forest analyses of the data are performed. For
each analysis, a relative importance score is calculated for each snp based on the minimum observed importance value.  Only 
those snp with large relative importance scores are determined to be in significant association with the trait.

\subsubsection{Single locus methods}
